A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-08 tidyindex.qmd
refine Section 2 to Tidy framework and bridge to pipeline

# Data pipeline {#sec-pipeline}

@@

Data pipeline refers to data transforming through a series of steps, similar to the concept composition of function in mathematics. To accomplish specific tasks, analysts initially write code snippets and later generalise them into functions. This generalization transforms the code from a specific use-case to an abstraction that is reuseable and easy to control with arguments. A pipeline represents a further generalization of functions into a workflow. It chains individual functions together, rather than creating a single monolithic pieces, to solve a data analysis problem. The resulting pipeline can be reused for other similar problems, and ideally, allows for easy switching of functions within the pipeline, similar to controlling function parameters through arguments.

<!-- One of the most commonly known pipeline examples is perhaps the Unix pipeline where programs can be concatenated with `|` to flow the output from the last program into the next program, i.e. ````command 1 | command 2 | command 3 | ...````. To solve a complex problem, the Unix system builds simple programs that do one thing well and work well together. -->

@@

Early discussion on data pipeline in R can be found in the interactive graphics literature [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014]. 
Specifically, @wickham_plumbing_2009 differentiates two different approaches used for structuring for (graphic) software: pipes and pipelines. The pipe approach offers functions, or components, and analysts concatenate functions together to form a pipeline to solve the specific problem at hands. This approach is used by the tidyverse suite, specifically in the dplyr package, where analysts apply functions in a sequence to address a wide array of data wrangling tasks. On the other hand, the pipeline approach addresses the data analysis tasks in a more structured manner, with a few centred steps used in almost all the workflow. For example, a supervised machine learning workflow typically involves steps like data splitting, model training and model evaluation, among others, regardless of the specific model used. A pipeline approach, like the tidymodel suite, modularizes these components and provides an unified syntax to specify different models.

@@

# A pipeline for building statistical indexes {#sec-index-pipeline}

@@

## Pipeline steps for constructing indexes {#sec-pipeline-steps}

@@

Tidyindex, as suggested by its name, follows the style and design principles of the tidyverse suite. The package is designed to simplify index construction for human analysts, ensure consistency across functions, and enable composability across modules to construct indexes. With the package, analysts can pass a data object through a set of pipeline modules introduced in @sec-pipeline-steps, offering different operations as standalone functions. For example, the dimension reduction module includes linear weighting, geometrical weighting, and manual expressions as possible methods to combine multivariate data. These methods are initially evaluated as a recipe of the operation without the data, and then, at the pipeline level, evaluating with the data using tidy evaluation, with the package `rlang` heavily used throughout the package.

>>>>

# Tidy framework {#sec-pipeline}

@@

The tidy framework consists of two key components: tidy data and tidy tools. The concept of tidy data [@wickham_tidy_2014] prescribes specific rules for organising data in an analysis, with observations as rows, variables as columns, and types of observational units as tables. Tidy tools, on the other hand, are concatenated to establish a sequential process through which the tidy data object flows, creating a pipeline for data processing. In this workflow, all the functions take inputs and outputs as tidy data, enabling direct concatenation of functions and saving analysts from bridging the outputs to subsequent operations. Additionally, the pipeline workflow breaks down complex problems into smaller, more manageable pieces. In an interactive data analysis, this flexibility to link various operations together as needed makes the pipeline workflow preferable over a monolithic design, where all the steps are predetermined and tightly integrated into a single piece.

@@

Dating back to earlier times, the pipeline approach to structure data analysis draws inspiration from the interactive graphics literature [@buja_elements_1988; @sutherland_orca_2000; @xie_reactive_2014; @wickham_plumbing_2009]. More recently, the tidy framework has provided a conceptual guide for the tidyverse suite [@wickham_welcome_2019] for general-purpose data wrangling, a concept that has gained popularity within the R community. Subsequently, more packages have since been developed to analyze data using the tidy framework for specific domains, for example, `tidytext` for text data [@silge_tidytext_2016], and `naniar` for missing data [@tierney_expanding_2023], and `tidymodels` for building machine learning models [@tidymodels]. To create a tidy workflow tailored to a specific domain, developers first need to identify the fundamental building blocks in the typical workflow. These components are then implemented as modules, which can be combined to form the pipeline. For example, in supervised machine learning models, steps such as data splitting, model training and model evaluation are commonly used in most workflow. In the `tidymodels`, these steps are correspondingly implemented as package `rsample`, `parsnip`, and `yardstick`, agnostic to the specific model chosen.

@@

# Details of the index pipline {#sec-index-pipeline}

@@

@@

### 2023-08-08 tidyindex.tex
refine Section 2 to Tidy framework and bridge to pipeline

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, breakable, boxrule=0pt, frame hidden, enhanced]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-pipeline}{%
\section{Data pipeline}\label{sec-pipeline}}

Data pipeline refers to data transforming through a series of steps,
similar to the concept composition of function in mathematics. To
accomplish specific tasks, analysts initially write code snippets and
later generalise them into functions. This generalization transforms the
code from a specific use-case to an abstraction that is reuseable and
easy to control with arguments. A pipeline represents a further
generalization of functions into a workflow. It chains individual
functions together, rather than creating a single monolithic pieces, to
solve a data analysis problem. The resulting pipeline can be reused for
other similar problems, and ideally, allows for easy switching of
functions within the pipeline, similar to controlling function
parameters through arguments.

Early discussion on data pipeline in R can be found in the interactive
graphics literature (Buja et al. 1988; Sutherland et al. 2000; Xie,
Hofmann, and Cheng 2014). Specifically, Wickham et al. (2009)
differentiates two different approaches used for structuring for
(graphic) software: pipes and pipelines. The pipe approach offers
functions, or components, and analysts concatenate functions together to
form a pipeline to solve the specific problem at hands. This approach is
used by the tidyverse suite, specifically in the dplyr package, where
analysts apply functions in a sequence to address a wide array of data
wrangling tasks. On the other hand, the pipeline approach addresses the
data analysis tasks in a more structured manner, with a few centred
steps used in almost all the workflow. For example, a supervised machine
learning workflow typically involves steps like data splitting, model
training and model evaluation, among others, regardless of the specific
model used. A pipeline approach, like the tidymodel suite, modularizes
these components and provides an unified syntax to specify different
models.

@@

\hypertarget{sec-index-pipeline}{%
\section{A pipeline for building statistical
indexes}\label{sec-index-pipeline}}

@@

\hypertarget{sec-pipeline-steps}{%
\subsection{Pipeline steps for constructing
indexes}\label{sec-pipeline-steps}}

@@

Tidyindex, as suggested by its name, follows the style and design
principles of the tidyverse suite. The package is designed to simplify
index construction for human analysts, ensure consistency across
functions, and enable composability across modules to construct indexes.
With the package, analysts can pass a data object through a set of
pipeline modules introduced in Section~\ref{sec-pipeline-steps},
offering different operations as standalone functions. For example, the
dimension reduction module includes linear weighting, geometrical
weighting, and manual expressions as possible methods to combine
multivariate data. These methods are initially evaluated as a recipe of
the operation without the data, and then, at the pipeline level,
evaluating with the data using tidy evaluation, with the package
\texttt{rlang} heavily used throughout the package.

@@

@@

@@

@@

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-pipeline}{%
\section{Tidy framework}\label{sec-pipeline}}

The tidy framework consists of two key components: tidy data and tidy
tools. The concept of tidy data (Wickham 2014) prescribes specific rules
for organising data in an analysis, with observations as rows, variables
as columns, and types of observational units as tables. Tidy tools, on
the other hand, are concatenated to establish a sequential process
through which the tidy data object flows, creating a pipeline for data
processing. In this workflow, all the functions take inputs and outputs
as tidy data, enabling direct concatenation of functions and saving
analysts from bridging the outputs to subsequent operations.
Additionally, the pipeline workflow breaks down complex problems into
smaller, more manageable pieces. In an interactive data analysis, this
flexibility to link various operations together as needed makes the
pipeline workflow preferable over a monolithic design, where all the
steps are predetermined and tightly integrated into a single piece.

Dating back to earlier times, the pipeline approach to structure data
analysis draws inspiration from the interactive graphics literature
(Buja et al. 1988; Sutherland et al. 2000; Xie, Hofmann, and Cheng 2014;
Wickham et al. 2009). More recently, the tidy framework has provided a
conceptual guide for the tidyverse suite (Wickham et al. 2019) for
general-purpose data wrangling, a concept that has gained popularity
within the R community. Subsequently, more packages have since been
developed to analyze data using the tidy framework for specific domains,
for example, \texttt{tidytext} for text data (Silge and Robinson 2016),
and \texttt{naniar} for missing data (Tierney and Cook 2023), and
\texttt{tidymodels} for building machine learning models (Kuhn and
Wickham 2020). To create a tidy workflow tailored to a specific domain,
developers first need to identify the fundamental building blocks in the
typical workflow. These components are then implemented as modules,
which can be combined to form the pipeline. For example, in supervised
machine learning models, steps such as data splitting, model training
and model evaluation are commonly used in most workflow. In the
\texttt{tidymodels}, these steps are correspondingly implemented as
package \texttt{rsample}, \texttt{parsnip}, and \texttt{yardstick},
agnostic to the specific model chosen.

@@

\hypertarget{sec-index-pipeline}{%
\section{Details of the index pipline}\label{sec-index-pipeline}}

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-tidymodels}{}}%
Kuhn, Max, and Hadley Wickham. 2020. \emph{Tidymodels: A Collection of
Packages for Modeling and Machine Learning Using Tidyverse Principles.}
\url{https://www.tidymodels.org}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-silge_tidytext_2016}{}}%
Silge, Julia, and David Robinson. 2016. {``Tidytext: {Text} {Mining} and
{Analysis} {Using} {Tidy} {Data} {Principles} in {R}.''} \emph{The
Journal of Open Source Software} 1 (3): 37.
\url{https://doi.org/10.21105/joss.00037}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-tierney_expanding_2023}{}}%
Tierney, Nicholas, and Dianne Cook. 2023. {``Expanding {Tidy} {Data}
{Principles} to {Facilitate} {Missing} {Data} {Exploration},
{Visualization} and {Assessment} of {Imputations}.''} \emph{Journal of
Statistical Software} 105 (February): 1--31.
\url{https://doi.org/10.18637/jss.v105.i07}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wickham_tidy_2014}{}}%
Wickham, Hadley. 2014. {``Tidy {Data}.''} \emph{Journal of Statistical
Software} 59 (September): 1--23.
\url{https://doi.org/10.18637/jss.v059.i10}.

\leavevmode\vadjust pre{\hypertarget{ref-wickham_welcome_2019}{}}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy
D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019.
{``Welcome to the {Tidyverse}.''} \emph{Journal of Open Source Software}
4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.
